Predicting computational reproducibility of data analysis pipelines \\in large population studies using collaborative filtering

### 2018-08-16 paper.tex
Updates following discussion with Soudabeh

Neuroinformatics pipelines are generally iterated on data coming from 10 to 
1,000 subjects, possibly with subtle input parameter variations to 
adjust specific data acquisition conditions. Subject data often capture 
anatomical and functional characteristics of their brain, for instance through
Magnetic Resonance Imaging (MRI) or Electroencephalography (EEG). The number of files 
associated with a subject may vary, and these files may also be of 
different sizes. Typical data processing times range from 15 minutes to 
15 hours per subject, with input ranging from 100~MB to 15~GB per subject, and 
outputs ranging from 1~GB to 500~GB.

The reproducibility of a given pipeline may vary across subjects, for 
instance due to different pipeline branches being executed depending on 
the nature, cardinality or content of the input data. For instance, in 
the public database released by the Human Connectome 
Project~\cite{van2013wu}, subjects may have one or two anatomical images of 
each modality; when two images are 
present, they are aligned together and averaged. The presence of 
artifacts, for instance due to high motion during the acquisition, may 
also trigger corrections that are not otherwise required. Some branches 
of the pipeline may also be executed only for specific acquisition 
parameters, for instance the intensity of the magnetic field. To 
capture all these variations, reproducibility evaluations need to be 
conducted on many subjects, which is unwieldy.

Our goal is to predict the outcome of reproducibility evaluations in a 
large population of subjects, from a reduced set of pipeline 
executions. More precisely, we aim at predicting whether a particular 
file produced by an analysis pipeline will be identical across 
execution conditions, or if it will contain reproducibility 
\emph{errors}. We approach the problem from the point of view of 
collaborative filtering, inspired by previous examples that applied 
this method outside of its initial application domain, for instance 
~\cite{feng2013efficient}. The main issue, and originality, of our 
problem lies in the fact that the training set cannot be arbitrarily 
sampled from the utility matrix. Instead, the training set has to 
respect the time constraints imposed by the order in which files 
are produced during pipeline executions.

This paper makes the following contributions:
\begin{enumerate}
\item We model reproducibility evaluations in data processing pipelines as a collaborative filtering problem.
\item We propose strategies to sample the training set under time constraints.
\item We evaluate and compare our sampling strategies on synthetic and real datasets.
\end{enumerate}
The problem formulation, collaborative filtering technique used, and 
proposed sampling strategies for the training set are presented in 
Section~\ref{sec:methods}. The datasets are described in 
Section~\ref{sec:datasets}, and experimental results are reported in 
Section~\ref{sec:results}.
We finally conclude on the best sampling method to use, and the 
potential impacts of the results, and on their limitations and 
generalizability.

@@

Collaborative filtering is a technique to predict unknown values of a 
matrix called ``utility matrix" from the known ones. Traditionally, the 
matrix represents the ratings of items, represented in columns, by 
users, represented in rows. Ratings might be explicit, when users 
provide ratings through a dedicated system, or implicit, when users' 
behaviors are tracked and analyzed to estimate their preferences. An 
overview of collaborative filtering is provided 
in~\cite{leskovec2014mining}. In our context, the utility matrix is the
pipeline matrix $U$ described previously.

@@

\begin{figure}[h!]
        \centering
        \begin{subfigure}[b]{0.45\columnwidth}
                  \includegraphics[width=\columnwidth]{figures/5vs7_columns_04_training}
                  \caption{Complete Columns}
                  \label{fig:Columns-Sample-Training-set}
        \end{subfigure}
        \begin{subfigure}[b]{0.45\columnwidth}
                  \includegraphics[width=\columnwidth]{figures/5vs7_rows_04_training}
                  \caption{Complete Rows}
                  \label{fig:Rows-Sample-Training-set}
        \end{subfigure}
        \begin{subfigure}[b]{0.45\columnwidth}
                 \includegraphics[width=\columnwidth]{figures/5vs7_random-real_04_training}
                \caption{Random Subjects (RS)}
                  \label{fig:Uniform-S-Sample-Training-set}
        \end{subfigure}
        \begin{subfigure}[b]{0.45\columnwidth}
                  \includegraphics[width=\columnwidth]{figures/5vs7_diagonal_04_training}
                  \caption{RFNU}
                  \label{fig:Diagonal-Sample-Training-set}
        \end{subfigure}
        \begin{subfigure}[b]{0.45\columnwidth}
                  \includegraphics[width=\columnwidth]{figures/5vs7_random-triangular-largest_04_training}
                  \caption{RFNTL}
                  \label{fig:triangular-L-Sample-Training-set}
        \end{subfigure}
        \begin{subfigure}[b]{0.45\columnwidth}
                  \includegraphics[width=\columnwidth]{figures/5vs7_random-triangular-smallest_04_training}
                  \caption{RFNTS}
                  \label{fig:triangular-S-Sample-Training-set}
        \end{subfigure}
        \begin{subfigure}[b]{\columnwidth}
        \centering
        \includegraphics[width=0.45\columnwidth]{figures/5vs7_random-unreal_04_training}
                  \caption{Random Unreal}
                  \label{fig:Random-Unreal-Sample-Training-set}
        \end{subfigure}
        \caption{Sample training set for different methods with a training ratio of 0.4.}
        \label{fig:sampling}
\end{figure}

\subsubsection{Random Subjects -- RS 
(Figure~\ref{fig:Uniform-S-Sample-Training-set})} This method builds 
the training set by selecting the files from random subjects until the 
training ratio is reached. The file selected in a subject is the file 
with the lowest index in this subject that has not been already 
selected in the training set. This respects the time constraints as files are 
sampled according to their production time-stamps.

@@

In each method, we also included the first row of the matrix (first 
generated file of each subject) and a random column (all files of a 
random subject) to avoid cold start issues.

@@

\subsection{Synthetic Data}

@@

\subsection{Real Data}

We collected data to evaluate the computational reproducibility of analysis
pipelines of the Human Connectome Project~\cite{glasser2013minimal}. We
processed a set S of 94 subjects randomly selected in the S500 HCP
release\footnote{\url{https://db.humanconnectome.org}} in three execution conditions with different
versions of the CentOS operating system (5.11, 6.8 and 7.2), using the
PreFreesurfer and Freesurfer pipelines described 
in~\cite{glasser2013minimal} and available on 
GitHub\footnote{\url{https://github.com/Washington-University/Pipelines/releases/tag/v3.19.0}}. 
For each pipeline, we identified the set F of files produced for all
subjects in all conditions. For each condition pair and each pipeline, 
we computed a binary reproducibility matrix U of size $|F|\times|S|$, 
where $U_{i,j}$ is true iif file $i$ of subject $j$ was different in 
each condition. Rows of $U$ were ordered by ascending file modification 
time in a random subject in S.

@@

\begin{figure}[h!]
  \centering
  \begin{subfigure}[b]{0.45\columnwidth}
        \includegraphics[width=\columnwidth]{data/Utility_Matrix/PreFreeSurfer/PFS_5v6_utility_matrix.png}
  \caption{PFS, C5 vs C6}
  \end{subfigure}
  \begin{subfigure}[b]{0.45\columnwidth}
         \includegraphics[width=\columnwidth]{data/Utility_Matrix/PreFreeSurfer/PFS_5v7_utility_matrix.png}
  \caption{PFS, C5 vs C7}
  \end{subfigure}
  \begin{subfigure}[b]{0.45\columnwidth}
        \includegraphics[width=\columnwidth]{data/Utility_Matrix/PreFreeSurfer/PFS_6v7_utility_matrix.png}
  \caption{PFS, C6 vs C7}
  \end{subfigure}
  \begin{subfigure}[b]{0.45\columnwidth}
        \includegraphics[width=\columnwidth]{data/Utility_Matrix/FS-100files/FS-First-bound-100files.png}
  \caption{FS (100 files), C6 vs C7}
  \end{subfigure}
  \caption{Utility matrices, PreFreesurfer (PFS) and Freesurfer (FS) pipelines}
    \label{fig:utility-matrices}
\end{figure}

@@

We conducted two experiments for each reproducibility matrix, to evaluate the performance of our 
predictions using (1) ALS without biases, and (2) ALS with biases. We 
compare the performance of our sampling methods to (1) a dummy 
classifier that always predicts the value in the majority class and (2) 
the Random Unreal method, used as the baseline sampling technique. All 
reported values are averages over 5 repetitions. Due to sampling 
issues, it is possible that the actual training ratio obtained with 
some of the sampling methods does not 
match the target one. We checked that the difference between the target 
and real training ratios was lower than 0.01. We used Spark's ALS model 
as available in package \texttt{pyspark.ml}, with 50 factors, a 
regularization parameter of 0.01, a maximum of 5 iterations and non 
negative constraints set to true.

@@

\subsection{Synthetic Data}

@@

However, only 3 methods are able to provide accuracy values above 0.85 
for all subject types: Random Subjects (RS), RFNU and RFNTL. This is 
due to the fact that these methods are the only ones that can sample 
files produced toward the end of the execution while maintaining a 
reduced number of empty columns. Surprisingly, RFNTS does not perform 
well for more than 2 subject types, even for a training ratio of 0.9.

@@

Figure~\ref{fig:results-synthetic-als-bias} shows accuracy results for 
ALS with bias. RS, RFNU and RFNTL remain the methods that best compare 
to Random Unreal, but the achieved accuracy is much lower than without 
biases. This is presumably due to 
the fact that file biases are all close to 0.5 in the synthetic data. 
In such a situation, biases are detrimental to the accuracy of the 
collaborative filtering process and shouldn't be included.

@@

\subsection{Real Data}

@@

Figure~\ref{fig:results-real-als} shows the accuracy results for ALS without bias. Among the methods that performed 
well in the synthetic dataset, RFNU performs the best, with an accuracy 
higher than 0.95 for all datasets, when the training ratio is larger 
than 0.5. RS and RFNTL are slightly underperforming compared to RFNU. 
Complete rows and complete columns do not even reach the accuracy of 
the dummy classifier. The reference method, Random Unreal, has an 
accuracy close to 1, which shows that collaborative filtering works 
well on this dataset too.

@@

Figure~\ref{fig:results-real-als-bias} shows the accuracy results for 
ALS with bias. From a training ratio of 0.5, all methods have a very 
good accuracy for all datasets, except complete columns in 
Figure~\ref{fig:pfs-c5vsc7-bias}. All methods except RFNU, RFNTL and 
RFNTS also perform well for a training ratio lower than 0.5. Overall, 
the accuracy of the methods is much higher than without bias, due to 
the fact that the file bias is very strong in this dataset, as 
it can be seen in Figure~\ref{fig:utility-matrices}.

@@

\begin{table}
\centering
\begin{tabular}{ccc}
& Sensitivity & Specificity \\
\hline
Complete Columns & 0.53 & 0.97\\
Complete Rows & 0.43 & 0.89\\
Random Subjects & 0.88 & 0.99\\
RFNU & 0.92 & 0.99 \\
RFNTL & 0.81 & 0.97\\
RFNTS & 0.65 & 0.93 \\
Random Unreal & 1 & 1
\end{tabular}
\caption{Average sensitivity and specificity values for the 6 synthetic datasets and Freesurfer.}
\label{table:roc}
\end{table}

\subsection{Varying the number of factors}
Figure~\ref{fig:factors} shows the effect of the number of factors used 
in the ALS optimization for the RFNU method, and the synthetic dataset with 8 types. 
For training ratios greater or equal than 0.5, the accuracy obtained 
with 3 factors is substantially higher than with 2 factors. Beyond 
3, the number of factors does not have an effect on the accuracy. 
This shows that the data is best explained by 3 factors. For training 
ratios lower than 0.5, the number of factors does not have any effect 
on the accuracy, presumably due to the low accuracy of the method.\todo{Soudabeh, check if the factors can be interpreted.}

\begin{figure}[h]
\includegraphics[width=0.9\columnwidth]{data/results/means_of_results/ALS/Synthetic/8type-RFNU-Differ-Num-Factors/ALS-8-types-ranks.pdf}
\caption{Effect of the number of factors used in ALS, RFNU method, synthetic dataset.}
\label{fig:factors}
\end{figure}

 \note{Soudabeh, add figure with two 3D plots showing the 3 factors of 
 every item and subject. Comment: the method indeed separates files in 
 3 blocks (for 8 types) and subjects in 8 types.}

\subsection{Varying the number of iterations}
Figure~\ref{fig:iterations} illustrates the effect of the number of 
iterations used in the ALS optimization, for the 
RFNU method, the synthetic dataset with 8 types, and 50 factors.

@@

\begin{figure}[h]
\includegraphics[width=0.9\columnwidth]{data/results/means_of_results/ALS/Synthetic/8type-RFNU-Differ-Num-Iterations-rank50/ALS-8-types-iteration.pdf}
\caption{Effect of the number of iterations used in ALS, RFNU method, 50 factors, synthetic dataset.}
\label{fig:iterations}
\end{figure}

\subsection{Error localization}

Figure~\ref{fig:overlap} compares the prediction errors made by RFNU, 
RFNTL and RFNTS. The matrices in this Figure represent the training set 
in black (negatives) and white (positives), and the test set in green 
(true positives), yellow (false negatives), gray (true negatives) and 
red (false positives). This representation provides insights regarding 
where, and perhaps why, prediction errors 
occur. At this training ratio, RFNU (Figure~\ref{fig:overlap-rfnu}) 
uniformly samples the number of files per subject between 0.8$N_f$ and 
$N_f$, which enables the training on the last 
files of the pipeline, while maintaining a low number of columns with 
a low training ratio. On the contrary, RFNTL 
(Figure~\ref{fig:overlap-rfntl}) samples the number of files per 
subject from 0.85$N_f$, but the probability to have a complete column 
is 0 (the one complete column in Figure~\ref{fig:overlap-rfntl} is the 
one included to prevent cold start issues), which leads to a ``stripe" 
of errors at the bottom of the matrix. RFNTS 
(Figure~\ref{fig:overlap-rfnts}) doesn't have this issue, as it 
includes many complete columns in the training set. However, it comes at 
the cost of several columns with a number of files lower than 
60\%: in such columns, the prediction is often entirely wrong, which explains
the reduced accuracy compared to RFNU.

\begin{figure*}
\begin{subfigure}[b]{0.6\columnwidth}
        \includegraphics[width=\columnwidth]{figures/RFNU_Syn_8_ALS_09_test_data_matrix_run1.png}
        \caption{RFNU: $U(0.8N_f, N_f)$}
        \label{fig:overlap-rfnu}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.6\columnwidth}
        \includegraphics[width=\columnwidth]{figures/RFNT-L_Syn_8_ALS_09_test_data_matrix_run2.png}
        \caption{RFNTL: $T(0.85N_f, 0.85N_f, N_f)$}
        \label{fig:overlap-rfntl}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.6\columnwidth}
        \includegraphics[width=\columnwidth]{figures/RFNT-S_Syn_8_ALS_09_test_data_matrix_run1.png}
        \caption{RFNTS: 0.3$T(0, N_f, N_f)$+0.7$N_f$}
        \label{fig:overlap-rfnts}
\end{subfigure}
\caption{Comparison between RFNU, RFNTL and RFNTS on synthetic dataset with 8 types, $\alpha=0.9$.}
\label{fig:overlap}
\end{figure*}

@@

From a practical standpoint, our study shows that reproducibility 
evaluations of the PreFreesurfer and Freesurfer pipelines can be 
conducted using only 50\% of the files produced by the evaluated 
pipelines, with an accuracy above 95\%. Potentially, this could reduce 
the computing time and storage required for such studies by a factor of 
2. However, such studies could not be conducted by processing only half 
of the subjects entirely, which would correspond to the Complete 
Columns sampling method. Instead, the processing of all the subjects 
should be initiated, and it should be terminated in a random uniform 
way, assuming that files are uniformly produced throughout the 
execution. On more complex matrices, such as the synthetic dataset 
studied here, the training ratio required to get a 95\% accuracy 
increases to 0.85 to 0.9.

@@

% Other possible examples
The variable of interest in our study was the reproducibility of files 
produced by the processing of a given subject, and it was controlled by 
a process, the evaluated pipeline. This process was hidden, since we 
considered the pipeline as a black box (we did not inspect its source 
code), it was deterministic, since it was implemented as a program, it 
was parametrized, as it behaved differently on different input datasets 
and parameters, and it was time-dependent, as the value of the 
variable of interest varied throughout the course of the pipeline 
execution. The method described in this paper could possibly be used to 
predict other variables controlled by such hidden, deterministic, 
parametrized, and time-dependent processes, for instance markers of 
chronic disease activity.

>>>>

Neuroinformatics pipelines are generally iterated on data coming from 10 to 
1,000 subjects, possibly with subtle input parameter variations to 
adjust specific data acquisition conditions. Subject data often capture 
anatomical and functional characteristics of their brain, for instance through
Magnetic Resonance Imaging (MRI) or Electroencephalography (EEG). The number of input files 
associated with a subject may vary, and these files may also be of 
different sizes. Typical data processing times range from 15 minutes to 
15 hours per subject, with input ranging from 100~MB to 15~GB per subject, and 
outputs ranging from 1~GB to 500~GB.

The reproducibility of a given pipeline may vary across subjects, for 
instance due to different pipeline branches being executed depending on 
the nature, cardinality or content of the input data. As an example, in 
the public database released by the Human Connectome 
Project~\cite{van2013wu}, subjects may have one or two anatomical images of 
each modality; when two images are 
present, they are aligned together and averaged. The presence of 
artifacts, for instance due to high motion during the acquisition, may 
also trigger corrections that are not otherwise required. Some branches 
of the pipeline may also be executed only for specific acquisition 
parameters, for instance the intensity of the magnetic field. To 
capture all these variations, reproducibility evaluations need to be 
conducted on many subjects, which is unwieldy.

Our goal is to predict the outcome of reproducibility evaluations in a 
large population of subjects, from a reduced set of pipeline 
executions. More precisely, we aim at predicting whether a particular 
file produced by an analysis pipeline will be identical across 
execution conditions, or if it will contain reproducibility 
\emph{errors}. We approach the problem from the point of view of 
collaborative filtering, inspired by works that applied 
this method outside of its initial application domain, for instance 
~\cite{feng2013efficient}. The main issue, and originality, of our 
problem lies in the fact that the training set cannot be arbitrarily 
sampled from the utility matrix. Instead, the training set has to 
respect the time constraints imposed by the file creation order during 
pipeline executions.

This paper makes the following contributions:
\begin{enumerate}
\item We model reproducibility evaluations in data processing pipelines as a collaborative filtering problem.
\item We propose strategies to sample the training set under time constraints.
\item We evaluate and compare our sampling strategies on synthetic and real datasets.
\end{enumerate}
The problem formulation, collaborative filtering technique used, and 
proposed sampling strategies for the training set are presented in 
Section~\ref{sec:methods}. The datasets are described in 
Section~\ref{sec:datasets}, and experimental results are reported in 
Section~\ref{sec:results}.
Finally, we conclude on the best sampling method to use, and on the 
impact, limitations and generalizability of the results.

@@

Collaborative filtering is a technique to predict unknown values of a 
matrix called ``utility matrix" from the known ones. Traditionally, the 
matrix represents the ratings of items, represented in columns, by 
users, represented in rows. Ratings might be explicit, when users 
provide ratings through a dedicated system, or implicit, when users' 
behaviors are analyzed to estimate their preferences. An 
overview of collaborative filtering is provided 
in~\cite{leskovec2014mining}. In our context, the utility matrix is the
matrix $U$ described previously.

@@

\begin{figure}[h!]
        \centering
        \begin{subfigure}[b]{0.45\columnwidth}
                  \includegraphics[width=\columnwidth]{figures/5vs7_columns_04_training}
                  \caption{Complete Columns}
                  \label{fig:Columns-Sample-Training-set}
        \end{subfigure}
        \begin{subfigure}[b]{0.45\columnwidth}
                  \includegraphics[width=\columnwidth]{figures/5vs7_rows_04_training}
                  \caption{Complete Rows}
                  \label{fig:Rows-Sample-Training-set}
        \end{subfigure}
        \begin{subfigure}[b]{0.45\columnwidth}
                 \includegraphics[width=\columnwidth]{figures/5vs7_random-real_04_training}
                \caption{Random Subjects (RS)}
                  \label{fig:Uniform-S-Sample-Training-set}
        \end{subfigure}
        \begin{subfigure}[b]{0.45\columnwidth}
                  \includegraphics[width=\columnwidth]{figures/5vs7_diagonal_04_training}
                  \caption{RFNU}
                  \label{fig:Diagonal-Sample-Training-set}
        \end{subfigure}
        \begin{subfigure}[b]{0.45\columnwidth}
                  \includegraphics[width=\columnwidth]{figures/5vs7_random-triangular-largest_04_training}
                  \caption{RFNTL}
                  \label{fig:triangular-L-Sample-Training-set}
        \end{subfigure}
        \begin{subfigure}[b]{0.45\columnwidth}
                  \includegraphics[width=\columnwidth]{figures/5vs7_random-triangular-smallest_04_training}
                  \caption{RFNTS}
                  \label{fig:triangular-S-Sample-Training-set}
        \end{subfigure}
        \begin{subfigure}[b]{\columnwidth}
        \centering
        \includegraphics[width=0.45\columnwidth]{figures/5vs7_random-unreal_04_training}
                  \caption{Random Unreal}
                  \label{fig:Random-Unreal-Sample-Training-set}
        \end{subfigure}
        \caption{Sample training set for different methods with a training ratio of 0.4 (dark elements are in the training set).}
        \label{fig:sampling}
\end{figure}

\subsubsection{Random Subjects -- RS 
(Figure~\ref{fig:Uniform-S-Sample-Training-set})} This method builds 
the training set by selecting the files from random subjects until the 
training ratio is reached. The file selected in a subject is the file 
with the lowest index in this subject that has not been already 
selected in the training set. This respects the time constraints as files are 
sampled according to their modification time-stamps.

@@

In each method, we also included the first row of the matrix (first 
 file of each subject) and a random column (all files of a 
random subject) to avoid cold start issues.

@@

\subsection{Synthetic Dataset}

@@

\subsection{Real Dataset}

We collected data to evaluate the computational reproducibility of analysis
pipelines of the Human Connectome Project~\cite{glasser2013minimal}. We
processed a set S of 94 subjects randomly selected in the S500 HCP 
release\footnote{\url{https://db.humanconnectome.org}} in three 
execution conditions with different versions of the CentOS operating 
system (5.11, 6.8 and 7.2 -- referred as C5, C6 and C7), using the 
PreFreesurfer and Freesurfer pipelines described 
in~\cite{glasser2013minimal} and available on 
GitHub\footnote{\url{https://github.com/Washington-University/Pipelines/releases/tag/v3.19.0}}. 
For each pipeline, we identified the set F of files produced for all
subjects in all conditions. For each condition pair and each pipeline, 
we computed a binary reproducibility matrix U of size $|F|\times|S|$, 
where $U_{i,j}$ is true if and only if file $i$ of subject $j$ was different in 
each condition. Rows of $U$ were ordered by ascending file modification 
time in a random subject in S.

@@

\begin{figure}[h!]
  \centering
  \begin{subfigure}[b]{0.45\columnwidth}
        \includegraphics[width=\columnwidth]{data/Utility_Matrix/PreFreeSurfer/PFS_5v6_utility_matrix.png}
  \caption{PFS, C5 vs C6}
  \end{subfigure}
  \begin{subfigure}[b]{0.45\columnwidth}
         \includegraphics[width=\columnwidth]{data/Utility_Matrix/PreFreeSurfer/PFS_5v7_utility_matrix.png}
  \caption{PFS, C5 vs C7}
  \end{subfigure}
  \begin{subfigure}[b]{0.45\columnwidth}
        \includegraphics[width=\columnwidth]{data/Utility_Matrix/PreFreeSurfer/PFS_6v7_utility_matrix.png}
  \caption{PFS, C6 vs C7}
  \end{subfigure}
  \begin{subfigure}[b]{0.45\columnwidth}
        \includegraphics[width=\columnwidth]{data/Utility_Matrix/FS-100files/FS-First-bound-100files.png}
  \caption{FS (100 files), C6 vs C7}
  \end{subfigure}
  \caption{Utility matrices for PreFreesurfer (PFS) and Freesurfer (FS). White cells denote reproducibility errors.}
    \label{fig:utility-matrices}
\end{figure}

@@

We conducted two experiments for each reproducibility matrix, to evaluate the performance of our 
predictions using (1) ALS without biases, and (2) ALS with subject and file biases. We 
compare the performance of our sampling methods to (1) a dummy 
classifier that always predicts the value in the majority class and (2) 
the Random Unreal method, used as the baseline sampling technique. All 
reported values are averages over 5 repetitions. Due to sampling 
issues, it is possible that the actual training ratio obtained with 
some of the sampling methods does not 
match the target one. We checked that the difference between the target 
and real training ratios was lower than 0.01. We used Spark's ALS model 
as available in package \texttt{pyspark.ml}, with 50 factors, a 
regularization parameter of 0.01, a maximum of 5 iterations and non 
negative constraints set to true.

@@

\subsection{Accuracy on Synthetic Data}

@@

However, only 3 methods are able to provide accuracy values above 0.85 
for all subject types: Random Subjects (RS), RFNU and RFNTL. 
Surprisingly, RFNTS does not perform well for more than 2 subject 
types, even for a training ratio of 0.9.

@@

Figure~\ref{fig:results-synthetic-als-bias} shows accuracy results for 
ALS with bias. RS, RFNU and RFNTL remain the methods that best compare 
to Random Unreal, but the achieved accuracy is much lower than without 
biases. This is presumably due to 
the fact that file biases are all close to 0.5 in the synthetic data. 
In such a situation, biases are detrimental to the accuracy of the 
collaborative filtering process and should not be included.

@@

\subsection{Accuracy on  Real Data}

@@

Figure~\ref{fig:results-real-als} shows the accuracy results for ALS without bias. Among the methods that performed 
well in the synthetic dataset, RFNU performs the best, with an accuracy 
higher than 0.95 for all datasets, when the training ratio is larger 
than 0.5. RS and RFNTL are underperforming compared to RFNU. 
Complete rows and complete columns do not even reach the accuracy of 
the dummy classifier. The reference method, Random Unreal, has an 
accuracy close to 1, which shows that collaborative filtering works 
well on this dataset too.

@@

Figure~\ref{fig:results-real-als-bias} shows the accuracy results for 
ALS with bias. From a training ratio of 0.5, all methods have a very 
good accuracy for all real datasets, except complete columns in 
Figure~\ref{fig:pfs-c5vsc7-bias}. All methods except RFNU, RFNTL and 
RFNTS also perform well for a training ratio lower than 0.5. Overall, 
the accuracy of the methods is much higher than without bias, due to 
the fact that the file bias is very strong in this dataset, as 
it can be seen in Figure~\ref{fig:utility-matrices}.

@@

\begin{table}
\centering
\begin{tabular}{ccc}
& Sensitivity & Specificity \\
\hline
Complete Columns & 0.53 & 0.97\\
Complete Rows & 0.43 & 0.89\\
Random Subjects & 0.88 & 0.99\\
RFNU & 0.92 & 0.99 \\
RFNTL & 0.81 & 0.97\\
RFNTS & 0.65 & 0.93 \\
Random Unreal & 1 & 1
\end{tabular}
\caption{Average sensitivity and specificity values for the 6 synthetic datasets and Freesurfer ($\alpha=0.9$).}
\label{table:roc}
\end{table}

\subsection{Effect of the number of factors}
Figure~\ref{fig:factors} shows the effect of the number of factors used 
in the ALS optimization for the RFNU method, and the synthetic dataset with 8 types. 
For training ratios greater than 0.5, the accuracy obtained 
with 3 factors is substantially higher than with 2 factors. Beyond 
3, the number of factors does not have an effect on the accuracy. 
This shows that the data is best explained by 3 factors. For training 
ratios lower than 0.5, the number of factors does not have any effect 
on the accuracy.

\begin{figure}[h]
\includegraphics[width=0.9\columnwidth]{data/results/means_of_results/ALS/Synthetic/8type-RFNU-Differ-Num-Factors/ALS-8-types-ranks.pdf}
\caption{Effect of the number of factors used in ALS, (RFNU, synthetic dataset, 8 types, 5 iterations).}
\label{fig:factors}
\end{figure}

Figure~\ref{fig:factors3d} confirms that 3 factors are indeed enough to 
separate the files of the synthetic dataset with 8 subject types in 3 
blocks, and the subjects in 8 types.

\begin{figure}[h]
\begin{subfigure}{\columnwidth}
\includegraphics[width=0.9\columnwidth]{./data/results/latent-factors/5i-3f-ALS-r1-08-userFactors.pdf}
\caption{Subject factors}
\end{subfigure}
\begin{subfigure}{\columnwidth}
\includegraphics[width=0.9\columnwidth]{./data/results/latent-factors/5i-3f-ALS-r1-08-itemFactors.pdf}
\caption{File factors}
\end{subfigure}
\caption{Subject and file factors produced by the collaborative filtering method (RFNU, synthetic dataset, 8 types, 3 factors, 5 iterations).}
\label{fig:factors3d}
\end{figure}

\subsection{Effect of the number of iterations}
Figure~\ref{fig:iterations} illustrates the effect of the number of 
iterations used in the ALS optimization, for the 
RFNU method, the synthetic dataset with 8 types, and 50 factors.

@@

\begin{figure}[h]
\includegraphics[width=0.9\columnwidth]{data/results/means_of_results/ALS/Synthetic/8type-RFNU-Differ-Num-Iterations-rank50/ALS-8-types-iteration.pdf}
\caption{Effect of the number of iterations used in ALS, (RFNU, synthetic dataset, 8 types, 50 factors).}
\label{fig:iterations}
\end{figure}

\subsection{Prediction error localization}

Figure~\ref{fig:overlap} compares the prediction errors made by RFNU, 
RFNTL and RFNTS. The matrices in this Figure represent the training set 
in black (negatives) and white (positives), and the test set in green 
(true positives), yellow (false negatives), gray (true negatives) and 
red (false positives). This representation provides insights regarding 
where, and perhaps why, prediction errors 
occur. At this training ratio, RFNU (Figure~\ref{fig:overlap-rfnu}) 
uniformly samples the number of files per subject between 0.8$N_f$ and 
$N_f$, which enables the training on the last 
files of the pipeline, while maintaining a low number of columns with 
a low training ratio. On the contrary, RFNTL 
(Figure~\ref{fig:overlap-rfntl}) samples the number of files per 
subject from 0.85$N_f$, but the probability to have a complete column 
is 0 (the one complete column in Figure~\ref{fig:overlap-rfntl} is the 
one included to prevent cold start issues), which leads to a ``stripe" 
of prediction errors at the bottom of the matrix. RFNTS 
(Figure~\ref{fig:overlap-rfnts}) doesn't have this issue, as it 
includes many complete columns in the training set. However, it comes at 
the cost of several columns with a number of files lower than 
60\%: in such columns, the prediction is often entirely wrong, which explains
the reduced accuracy compared to RFNU.

\begin{figure*}
\begin{subfigure}[b]{0.5\columnwidth}
        \includegraphics[width=\columnwidth]{figures/RFNU_Syn_8_ALS_09_test_data_matrix_run1.png}
        \caption{RFNU: $U(0.8N_f, N_f)$}
        \label{fig:overlap-rfnu}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.5\columnwidth}
        \includegraphics[width=\columnwidth]{figures/RFNT-L_Syn_8_ALS_09_test_data_matrix_run2.png}
        \caption{RFNTL: $T(0.85N_f, 0.85N_f, N_f)$}
        \label{fig:overlap-rfntl}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.5\columnwidth}
        \includegraphics[width=\columnwidth]{figures/RFNT-S_Syn_8_ALS_09_test_data_matrix_run1.png}
        \caption{RFNTS: 0.3$T(0, N_f, N_f)$+0.7$N_f$}
        \label{fig:overlap-rfnts}
\end{subfigure}
\caption{Comparison between RFNU, RFNTL and RFNTS on synthetic dataset with 8 types, $\alpha=0.9$.}
\label{fig:overlap}
\end{figure*}

@@

From a practical standpoint, our study shows that reproducibility 
evaluations of the PreFreesurfer and Freesurfer pipelines can be 
conducted using only 50\% of the files produced by the evaluated 
pipelines, with an accuracy above 95\%. Potentially, this could reduce 
the computing time and storage required for such studies by a factor of 
2. However, such studies could not be conducted by processing only half 
of the subjects entirely, which would correspond to the Complete 
Columns sampling method. Instead, the processing of all the subjects 
should be initiated, and it should be terminated in a random uniform 
way, assuming that files are uniformly produced throughout the 
execution. On more complex matrices, such as the synthetic dataset 
studied here, the training ratio required to get a 95\% accuracy 
increases to 0.85.

@@

% Other possible examples
The variable of interest in our study was the reproducibility of files 
produced by the processing of a given subject. This variable was controlled by 
a process, the evaluated pipeline. This process was hidden, since we 
we did not inspect the pipeline source 
code, it was deterministic, since it was implemented as a program, it 
was parametrized, as it behaved differently on different input datasets 
and parameters, and it was time-dependent, as the value of the 
variable of interest varied throughout the course of the pipeline 
execution. The method described in this paper could possibly be used to 
predict other variables controlled by such hidden, deterministic, 
parametrized, and time-dependent processes, for instance markers of 
chronic disease activity.
